Retrofitting Effect Handlers onto OCaml

### 2020-11-10 retro-concurrency.tex
Added related work

\begin{figure*}
\begin{minipage}{0.35\linewidth}
  \centering
  \includegraphics[scale=0.45]{figures/fiber}
  \subcaption{Fiber layout}
  \label{fig:fiber}
\end{minipage}
\begin{minipage}{0.64\linewidth}
  \begin{minipage}{\linewidth}
    \centering
    \includegraphics[scale=0.4]{figures/multicore_stack}
    \subcaption{Stack layout for \texttt{meander} example from \S\ref{sec:stack}.}
    \label{fig:mcstack}
    \vspace{2mm}
  \end{minipage}
  %
  \begin{minipage}{0.55\linewidth}
    \begin{lstlisting}
effect E : unit
effect F : unit;;
match (* comp_e *)
  match (* comp_f *)
    (*p1*) perform E (*p3*)
  with | v -> v | effect F kf -> ...
with | v -> v
		 | effect E ke ->
		     (*p2*) continue ke ()
    \end{lstlisting}
		\subcaption{Constructing continuation objects}
		\label{code:effimpl}
  \end{minipage}
  %
  \begin{minipage}{0.44\linewidth}
    \centering
    \includegraphics[scale=0.42]{figures/fiber_handler}
		\subcaption{Program state for code in~\ref{code:effimpl}}
    \label{fig:fiber_handler}
  \end{minipage}
\end{minipage}
\vspace{2mm}
\mycaption{Layout of Multicore OCaml effect handlers.}
\end{figure*}

@@

Finally, we have the saved exception pointer, which points to the top-most
exception frame, and the saved stack pointer, which points to the top of the
stack. Switching between fibers only involves saving the exception and the
stack pointer of the current stack and loading the same on the target stack.
Since OCaml does not generate pointers into the stack, the two |fiber_info|
fields are the only ones that need to be updated when fibers are moved. We
allocate the space for the fiber using |malloc| and |free| it when no longer
required. We do not use a \emph{stack cache} of recently freed stacks for
speeding up fiber allocation~\ref{Farvardin}. In our experiments, we observed
that directly using |mimalloc| outperforms having the stack cache.

@@

\begin{figure}
	\begin{minipage}{0.49\linewidth}
		\centering
		\includegraphics[width=\linewidth]{http-benchmarks/throughput.pdf}
		\subcaption{Throughput}
		\label{grf:throughput}
	\end{minipage}
	\begin{minipage}{0.49\linewidth}
		\includegraphics[width=\linewidth]{http-benchmarks/latency.pdf}
		\subcaption{Tail latency}
		\label{grf:latency}
	\end{minipage}
	\mycaption{Web server performance.}
	\vspace{-2mm}
\end{figure}

@@

\section{Conclusions and Future Work} \label{sec:conc}

>>>>

\begin{figure*}
\begin{minipage}{0.35\linewidth}
  \centering
  \includegraphics[scale=0.45]{figures/fiber}
  \subcaption{Fiber layout}
  \label{fig:fiber}
\end{minipage}
\begin{minipage}{0.64\linewidth}
  \begin{minipage}{\linewidth}
    \centering
    \includegraphics[scale=0.4]{figures/multicore_stack}
    \subcaption{Stack layout for \texttt{meander} example from \S\ref{sec:stack}.}
    \label{fig:mcstack}
    \vspace{2mm}
  \end{minipage}
  %
  \begin{minipage}{0.55\linewidth}
    \begin{lstlisting}
effect E : unit
effect F : unit;;
match (* comp_e *)
  match (* comp_f *)
    (*p1*) perform E (*p3*)
  with | v -> v | effect F kf -> ...
with | v -> v
		 | effect E ke ->
		     (*p2*) continue ke ()
    \end{lstlisting}
		\subcaption{Constructing continuation objects}
		\label{code:effimpl}
  \end{minipage}
  %
  \begin{minipage}{0.44\linewidth}
    \centering
    \includegraphics[scale=0.42]{figures/fiber_handler}
		\subcaption{Program state for code in~\ref{code:effimpl}}
    \label{fig:fiber_handler}
  \end{minipage}
\end{minipage}
\mycaption{Layout of Multicore OCaml effect handlers.}
\end{figure*}

@@

Finally, we have the saved exception pointer, which points to the top-most
exception frame, and the saved stack pointer, which points to the top of the
stack. Switching between fibers only involves saving the exception and the
stack pointer of the current stack and loading the same on the target stack.
Since OCaml does not generate pointers into the stack, the two |fiber_info|
fields are the only ones that need to be updated when fibers are moved. We
allocate the space for the fiber using |malloc| and |free| it when no longer
required. We do not use a \emph{stack cache} of recently freed stacks for
speeding up fiber allocation~\ref{Farvardin}. In our experiments, we observed
that directly using |mimalloc|~\cite{Leijen19} outperforms having the stack
cache.

@@

\begin{figure}
	\begin{minipage}{0.49\linewidth}
		\centering
		\includegraphics[width=\linewidth]{http-benchmarks/throughput.pdf}
		\subcaption{Throughput}
		\label{grf:throughput}
	\end{minipage}
	\begin{minipage}{0.49\linewidth}
		\includegraphics[width=\linewidth]{http-benchmarks/latency.pdf}
		\subcaption{Tail latency}
		\label{grf:latency}
	\end{minipage}
	\mycaption{Web server performance.}
	\vspace{-4mm}
\end{figure}

@@

There are several strategies for implementing effect handlers. Eff~\cite{Eff},
Helium~\cite{Biernacki20}, Frank~\cite{Frank}, Links server
backend~\cite{Hillerstrom20} use an interpreter similar to our operational
semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
selective CPS translation. These language are equipped with an effect system,
which allows compiling pure code in direct-style and effectful code in
continuation passing style. Leijen et al.~\cite{Leijen14} implements effect
handlers as a library in C using stack copying. Since C allows pointers into
the stack, care is taken to ensure that when continuations are resumed, the
constituent frames are restored to the same memory addresses as at the time of
capture. Kiselyov et al.~\cite{Kiselyov18} use an implementation of
multi-prompt delimited continuations as an OCaml library~\cite{Kiselyov12} to
embed Eff language in OCaml. Indeed, Forester et al.~\cite{Forster19} showed
that in an untyped setting, effect handlers, monadic reflection and delimited
control can macro-express each other.

Multicore OCaml effect handlers are similar to~\cite{Leijen14, Kiselyov12}, but
are more efficient; since we only allow our continuations to be resumed at-most
once, there is no need for copying frames. Bruggeman et al.~\cite{Bruggeman96}
show how to implement one-shot continuations efficiently using segmented stacks
in Scheme. Farvardin et al.~\cite{Farvardin20} perform a comprehensive
evaluation of various implementation strategies for continuations on modern
hardware. Multicore OCaml stacks do not neatly fit the description of one of
these implementation strategies. They best described as using the resize
strategy from Farvardin et al. for each of the fibers, which are in turn linked
to represent the current stack and the captured continuations.

Kawahara et al.~\cite{Kawahara20} implement one-shot effect handlers using
coroutines (a non-local control flow mechanism) as a macro-expressible
translation, and present an embedding in Lua and Ruby. Lua provides asymmetric
coroutines~\cite{Lua} where each coroutine uses its own stack similar to how
each handled computation runs in its own fiber in Multicore OCaml. Similar to
Multicore OCaml, Lua does not allow coroutines to include C frames. It is
useful to note that Lua is an interpreted language unlike OCaml and DWARF
unwind tables are not a concern in Lua.

The interaction on non-local control flow and resources have been studied
extensively. Scheme uses |dynamic-wind| to ensure de-allocation (and
re-allocation) of resources when the non-local control leaves and enters back
into a context. |dynamic-wind| is known to be complex and not quite what one
wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main challenge is distinguishing
non-returning exits from ones that would, and clean up to occur only on
non-returning exit. In Multicore OCaml, we assume that continuations are used
linearly, and hence, no clean up occurs when effects are performed but only when
an exception is thrown, perhaps from the use of |discontinue|. This obviates
the need for |dynamic-wind|. Leijen~\cite{Leijen18} explicitly extends effect
handlers with |finally| clause. Instead, Multicore OCaml |discontinue|
primitive builds on extant defensive coding practices against exceptions to
clean up resource.

Languages with effect handlers typically allow continuations to be resumed more
than once to enable use cases such as backtracking search and probabilistic
programming~\cite{Pyro}. Multicore OCaml has an experimental feature for
explicitly cloning continuations, which makes a copy of each of its constituent
fibers. Multi-shot continuations are difficult to reason about in the presence
of linear resources such as file descriptors. Recall that OCaml programs are
written with the expectation that function calls return \emph{exactly-once}
either normally or exceptionally. With multi-shot continuations, they may
return more than once, leading to errors such as double-freeing malloc'ed
memory. Since our goal is to retrofit effect handlers onto OCaml, we relegate
continuation cloning to the OCaml |Obj| module, which contains unsafe features.

\section{Conclusions}
\label{sec:conc}

We presented the motivation, design and implementation of effect handlers
specifically aimed at retrofitting them onto OCaml with tool compatibility,
negligible overheads for existing OCaml programs and good performance for those
that use effect handlers. Our evaluation shows that this is indeed the case.
